---
title: "Phyloseq PreProcessing"
author: "sophiaaredas"
date: "`r Sys.Date()`"
output: html_document
  toc: yes
  toc_float:
    collapsed: no
    smooth_scroll: yes
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---
```{r}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../figures/02_PreProcessing/") #send any figure output to this folder
```


# Load libraries
```{r load-libraries}
library(phyloseq)
library(devtools)
library(tidyverse)
```

# Goals
Here we will process the data nito a phyloseq object.
- ASV table
- taxonomy table
- track reads (metadata)

Then, we will remove the following:
1. mitochondria 
2. chloroplast
3. remove samples without "enough" reads 

Finally, write data file of phyloseq output

# Load data

## ASV Table
```{r load-asv-table}
#First load asv table
load("/local/workdir/sna49/git_repos/SalinityGradient_16S/data/01_DADA2/ASV_counts.RData")

#Inspect asv_tab
head(asv_tab)[,1:5]

#fix names
sample_names <- colnames(asv_tab)
samples_fixed <- sapply(strsplit(basename(sample_names), "_"), `[`,1) 
head(samples_fixed)

colnames(asv_tab) <- samples_fixed
```


## Taxonomy table
```{r load-tax-table}
tax_df <- read.table("/local/workdir/sna49/git_repos/SalinityGradient_16S/data/01_DADA2/ASV_taxonomy.tsv", sep = "\t", skip = 1)
head(tax_df)

#fix column names
colnames(tax_df) <- c("asv_names", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "ASV", "ASVseq")
head(tax_df)

#taxonomy table matrix 
tax_mat <-
  tax_df %>% 
    tibble::column_to_rownames(., var = "asv_names") %>% 
    as.matrix()
```

## Track reads data
```{r load-track-reads}
load("/local/workdir/sna49/git_repos/SalinityGradient_16S/data/01_DADA2/track_read_counts.RData")

#take a look at data
track_counts_df
dim(track_counts_df)
# [1] 95  8
#95 rows, 8 columns

#in termianl copy metadata.csv into Salinity_Gradient_16S/data folder

#from: /workdir/in_class_data/SalinityGradient_16S/
# cp metadata.csv /local/workdir/sna49/git_repos/SalinityGradient_16S/data/01_DADA2

# load in metadata
metadata_df <- read.csv("/local/workdir/sna49/git_repos/SalinityGradient_16S/data/metadata.csv")
dim(metadata_df)
colnames(metadata_df)

# merge metadata_df with track_reads_df
metadata_track_reads_df <- 
  metadata_df %>% 
  left_join(., track_counts_df, by = "names") %>% 
  dplyr::select(-X)

#intutition check
head(metadata_track_reads_df)

#update row.names to be sample names
#before
row.names(metadata_track_reads_df)
#rewrite
row.names(metadata_track_reads_df) <- metadata_track_reads_df$names
#check afterwards taht it worked
row.names(metadata_track_reads_df)

#intution check
head(metadata_track_reads_df)
```

## Handoff to phyloseq
```{r phyloseq-handoff}
#double check its all good
dim(asv_tab)
dim(tax_mat)
#these two match 3080 so yay

#intutition with base r
stopifnot(row.names(asv_tab) == row.names(tax_mat))
#theres no output which means its happy

#construct the phyloseq object
raw_physeq <- phyloseq(otu_table(asv_tab, taxa_are_rows = TRUE), 
                       sample_data(metadata_track_reads_df), 
                       tax_table(tax_mat))

#save this raw phyloseq object
save(raw_physeq, file = "/local/workdir/sna49/git_repos/SalinityGradient_16S/data/02_PreProcessing/raw_physeq.RData")
```

# Clean up the data
remove:
1. chloroplasts
2. mitochondria
```{r rm-mitos-chloros}
#remind myself of tax table
View(tax_mat)
#make new physeq without chloroplasts
noChloros_physeq <- raw_physeq %>% 
  #rm chloroplasts
  subset_taxa(Order != "Chloroplast" | is.na(Order))

#how many taxa were chloroplasts
num_chloro_ASVs <- ntaxa(raw_physeq) - ntaxa(noChloros_physeq)

noChloros_physeq %>% 
  tax_table() %>% 
  data.frame() %>% 
  View()
  
#remove mitochondria
noChlorosMitos_physeq <- noChloros_physeq %>% 
    subset_taxa(Family != "Mitochondria" | is.na(Family))

#how many mitochondrial ASVs?
num_mito_ASVs <- ntaxa(noChlorosMitos_physeq) - ntaxa(noChlorosMitos_physeq)
num_mito_ASVs 

noChlorosMitos_physeq

#how many total asvs were removed from chloros and mitos
ntaxa(raw_physeq) - ntaxa(noChlorosMitos_physeq)

#proportion of asvs?
ntaxa(noChlorosMitos_physeq)/ntaxa(raw_physeq)
```

# Evaluate and remove the control samples

Take a look at negative controls and make a decision about whether or not to remove the ASVs that we found in our controls

1. negative controls
2. ASVs found within the negative controls and their distribution in teh samples
3. evaluate the mock community

## Negative controls
```{r neg-controls}
# create a vector of samples that were negative controls
control_samples <- c("WaterControl", "022um-Control", "3um-Control", "DNA-Ext-Control")

# make a new phyloseq
control_physeq <- noChlorosMitos_physeq %>% 
  subset_samples(., names %in% control_samples) %>% 
  prune_taxa(taxa_sums(.) > 0, .)

control_physeq

#vector of control asvs
control_ASVs <-
  control_physeq %>% 
  tax_table() %>% 
  data.frame() %>% 
  dplyr::select(ASV) %>% 
  as.vector()

#lets visualize raw abundance values of control ASVs
control_physeq %>% 
  plot_bar(., "ASV", fill = "ASV")
#so we see that ASV3 is super highly abundant so we will check the control ASVs

######## Check the control ASVs in teh samples
#make new phyloseq object with only the control ASVs and ALL samplse
controlASV_physeq <- noChlorosMitos_physeq %>% 
  subset_taxa(., ASV %in% control_ASVs$ASV)

#inspect it
controlASV_physeq

#lets take a look at abundances
controlASV_physeq %>% 
  plot_bar(., "ASV", fill="ASV", facet_grid = Sample_or_Control~.)

# check tax of ASV_3
# controlASV_physeq %>% 
#   tax_table() %>% 
#   data.frame() %>% 
#   View()

# we can get rid fo ASV3 but keep ASV18 and 56. ASV3 is Pseudomonas which is likely contamination 

# would only like to keep ASV_18 and ASV56. everything else is trash
controlASVs_toREMOVE <- dplyr::setdiff(control_ASVs$ASV, c("ASV_18", "ASV_56"))

#double check that its 29 as output
length(controlASVs_toREMOVE)

## make new phyloseq object
noControlsChlorosMitos_physeq <- 
  noChlorosMitos_physeq %>% 
  # actually remove the cotrol ASVs except 18 and 56
  subset_taxa(., !(ASV %in% controlASVs_toREMOVE)) %>% 
  #remove control samples
  subset_samples(., !(names %in% control_samples))

#inspect no controls object
noControlsChlorosMitos_physeq

#asfter filtering how many ASVs were removed
total_ASVs_rm <- ntaxa(raw_physeq) - ntaxa(noControlsChlorosMitos_physeq)

#output should be 395 of ASVs that were removed 
```

